---
title: "multilevelcoda data generation"
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(extraoperators)
library(compositions)
library(multilevelcoda)
library(brms)
library(cmdstanr)
library(insight)

library(doFuture)
library(foreach)
library(parallel)
library(MASS)
```

\newpage

# Inputs 

```{r}
meanscovs <- readRDS("meanscovs.RDS") # mean and covariance to generate data
source("input.R") # conditions and functions to generate data and run models

# sim model script
# source("simmodel.R")
```

\newpage

# Data generation process
\textcolor{red}{The model we are using is a multilevel model with compositional predictors, 
specifically daily $D$-part sleep-wake behaviours predicting sleepiness. 
It corresponds to the Multilevel models with compositional predictors section in the package paper.
In this study, model includes random intercept, no random slopes}.

## Mathematical specification
To express a model for the time-varying $D$-part multilevel compositional predictor, 
we first denote the repeated measure observed at time point $i$ for individual $j$ as $y_{ij}$. 
The prediction of the outcome $y_{ij}$ is the linear combination of the between-person and 
within-person effects of a $D$-part composition (expressed as a set of $D-1$-part or $ilr$ coordinates). 
We write

$$
y_{ij} = \beta_{0j} + 
  \overbrace{\sum_{k = 1}^{D-1} \beta_k bz_{jk}}^{\text{between}} + 
  \underbrace{\sum_{k = 1}^{D-1}\beta_{(k + D - 1)j}\hspace{0.1cm} wz_{ijk}}_{\text{within}} +
  \varepsilon_{ij}
$$

where 
${y}_{ij}$ a repeated measure observed at time point $i$ for individual $j$. 
This was generated through the linear combination of the between-person and 
within-person effects of a $D$-part composition (expressed as a set of $(D - 1)$-part or $ilr$ coordinates).

with population level parameters: 
$$\beta_{0j} = \gamma_{00} + u_{0j}$$

and
$$
\begin{bmatrix}
\beta_{1} \\
\beta_{2} \\
\vdots \\
\beta_{(k + D - 1)}
\end{bmatrix} 
= 
\begin{bmatrix}
\gamma_{01} \\
\gamma_{02} \\
\vdots \\
\gamma_{(k + D - 1)}
\end{bmatrix}
$$
person level random effects: 

$$u_{0j} \sim \mathcal{N}(0, \sigma^{2}_{u})$$


and within-person random error:

$$\varepsilon_{ij} \sim \mathcal{N}(0, \sigma^{2}_{\varepsilon})$$

### Model specification for a 5-part compositional predictor

$$
y_{ij} = \beta_{0j} + 
  \overbrace{\sum_{k = 1}^{4} \beta_k bz_{jk}}^{\text{between}} + 
  \underbrace{\sum_{k = 1}^{4}\beta_{(k + 4)}\hspace{0.1cm} wz_{ijk}}_{\text{within}} +
  \varepsilon_{ij}
$$

with: 
$$\beta_{0j} = \gamma_{00} + u_{0j}$$

and
$$
\begin{bmatrix}
\beta_{1} \\
\beta_{2} \\
\vdots \\
\beta_{8}
\end{bmatrix} 
= 
\begin{bmatrix}
\gamma_{01} \\
\gamma_{02} \\
\vdots \\
\gamma_{08} 
\end{bmatrix}
$$

with 

$$u_{0j} \sim \mathcal{N}(0, \sigma^{2}_{u})$$


and

$$\varepsilon_{ij} \sim \mathcal{N}(0, \sigma^{2}_{\varepsilon})$$
The between- and within-person components of the composition are $bz$ and $wz$, with 
the subscripts denoting that the between component is unique to individual $j$ 
and the within component is unique to time $i$ for individual $j$. 
The between- and within-person components for the 
relationship between composition and the outcome are $\beta_k$ and $\beta_{k + D - 1}$
The subscript $k$ indicates the index of the $\beta$ for the composition predictor 
(expressed as a set of $ilr$ coordinates).
Because each $ilr$ coordinate is decomposed into its between- and within-person components, 
the total number of the $\beta$ for the $ilr$ coordinates is double the number of the $ilr$ coordinates, 
hence the subscript of the $\beta$ for $wz$ is $k + D - 1$.

\newpage

## Code specification

```{r, eval = FALSE}

library("mvtnorm")
library(MASS)

n <- c(30, 50, 360, 1200)   # number of people
k <- c(3, 5, 7, 14)     # number of time points per person
Ddash <- c(2, 3, 4) # number of ilr coordinates
rint_sd <- c(1, sqrt(.5), sqrt(1.5)) # levels of random effect
res_sd <- c(1, sqrt(.5), sqrt(2)) # levels of residual error

# FOR 5-PART COMPOSITION
gamma_00 <- groundtruth$b_Intercept # some Ddash length vector of ilr population intercepts

gamma_01 <- groundtruth$b_bilr1
gamma_02 <- groundtruth$b_bilr2 
gamma_03 <- groundtruth$b_bilr3 
gamma_04 <- groundtruth$b_bilr4
gamma_05 <- groundtruth$b_wilr1 
gamma_06 <- groundtruth$b_wilr2
gamma_07 <- groundtruth$b_wilr3
gamma_08 <- groundtruth$b_wilr4

# x_ij <- ...     # n*m covariates

# 0_Ddash <- 0 # mean vector of random effects and error

Omega_w  <- res_sd # some Ddash x Ddash matrix of variance covariance
Omega_0b <- rint_sd # some Ddash x Ddash matrix of variance covariance
# Omega_1b <- ... # don't have random slope in the model atm :D

u_0j <- 
  rnorm(
    n = n, 
    mean = 0,
    sd = Omega_0b
  )
  
# u_1j <- 
#   rmvnorm(
#     n = n, 
#     mean = 0,
#     sigma = Omega_1b
#   )

e_ij <- 
  rnorm(
    n = n * k, 
    mean = 0,
    sd = Omega_w
  )

# y is sleepy
# a linear combination of gamma_00, gamma_01, u_0j, u_1j, e_ij
y <- gamma_00  +
  (gamma_01 * bilr1) +
  (gamma_02 * bilr2) +
  (gamma_03 * bilr3) +
  (gamma_04 * bilr4) +
  (gamma_05 * wilr1) +
  (gamma_06 * wilr2) +
  (gamma_07 * wilr3) +
  (gamma_08 * wilr4) +
  u_0j + e_ij
  
# some other inputs for calculating ILRs
sbp <- matrix(c(
  1, 1, -1,-1, -1,
  1, -1, 0, 0, 0,
  0, 0, 1, -1, -1,
  0, 0, 0, 1, -1), ncol = 5, byrow = TRUE) 
# psi <- gsi.buildilrBase(t(sbp))
parts <- c("TST", "WAKE", "MVPA", "LPA", "SB")
```

\newpage
# Model specification in `R/brms`

```{r, eval = FALSE}

head(dataset, n = 10)

# the "q" below allows the random intercepts to be correlated over outcome vars
# e.g., `cor(ilr1, ilr2)` for the random intercepts within person id 
bf1 <- brm(sleepy ~ bilr1 + bilr2 + bilr3 + bilr4 + wilr1 + wilr2 + wilr3 + wilr4 + (1 | ID),
           data = dataset,
           cores = 4,
           chains = 4,
           iter = 3000,
           warmup = 500,
           backend = "cmdstanr")

# or using multilevelcoda
cilr <- compilr(dataset, sbp, parts, total = 1440, idvar = "ID")

bf2 <- brmcoda(
  cilr,
  sleepy ~ bilr1 + bilr2 + bilr3 + bilr4 + wilr1 + wilr2 + wilr3 + wilr4 + (1 | ID),
  cores = 4,
  chains = 4,
  iter = 3000,
  warmup = 500,
  backend = "cmdstanr")

```

\newpage

# A note on the response vector's distribution

The distribution of the response vector $\boldsymbol{z}_{ij}^T$ is determined purely from the right hand side of the model equation. As the right hand side of the model equation is a linear combination of multivariate normal random variables, the response vector for person $j$ at time point $i$ is also multivariate normal distributed with mean corresponding to the fixed effects and variance corresponding to the random effects (and error):

$$
\boldsymbol{z}_{ij}
\sim 
\mathcal{N}\left(
\boldsymbol{\gamma}_{00} + \boldsymbol{\gamma}_{01},
\Omega_{0B} + x_{ij} \Omega_{1B} + \Omega_W 
\right)
$$










